Detecting confounder and center bias in machine learning models

### 2021-10-28 manuscript.tex
Updates from Overleaf

Predictive modelling and supervised learning methods have recently become increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized, spisak2020pain, walsh2021dome}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Biomedical validity (i.e. whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers \citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases \citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability. Such biases can originate form many sources.

First, measurement artifacts can obviously be considered as a potential source of confounder bias. For instance, magnetic resonance imaging (MRI)-based predictive models can be driven by in-scanner head motion artifacts, next to (or instead of) signal of neural origin, especially, if the prediction target is also associated with motor function and motion characteristics, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive}, attention deficit hyperactivity disorder (ADHD) \citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD) \citep{gotts2013perils, spisak2014voxel, spisak2019optimal}. Likewise, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.

Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling. For instance, resting-state functional connectivity has been shown to be strongly predictive to age \citep{dukart2011age, wang2012decoding} and moderately to fluid intelligence \citep{ cole2012global, he2020deep}. As fluid intelligence is known to decline with aging \citep{kievit2018neural}, models trained to predict intelligence might provide a statistically significant predictive performance by picking up solely on age-related variance \citep{dubois2018distributed, lohmann2021predicting}.
Importantly, the target variable must not necessarily display a true association to the confoudner to produce a biased model. Model training can be sensitive to sampling bias and stochastic group differences in the training set; a phenomenon that might have calamitous consequences, as increasingly recognised, for instance, in case of racially biased machine learning models \citep{ obermeyer2019dissecting, lwowski2021risk}.

Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target. Such models are hard to calibrate, might output systematically biased predictions and might display a worse-than-expected performance when applied on data from new centers/batches.

Various data cleansing, correction and harmonization approaches have been proposed to mitigate detrimental effects of confounders, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}. However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.

Powerful and robust statistical tests to quantify confounding effects in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches. Such methods, however, have to tackle the fact that the conditional distribution of the models' output, given the target, is often non-normal and non-linear, as a consequence of e.g. model regularization \citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.

Recently, three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder effects in predictive modelling.

The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via 'restricted permutations' that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature). However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances. Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control in certain, realistic scenarios.
The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, on the other hand, does not provide any quantification of statistical significance.
Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high a computational cost (especially when hyperparameters are optimized in a nested cross-validation).
The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there is an unobserved random variable that is the cause of both. However, this null hypothesis might not be of interest in many real applications. For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation. In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.

@@

In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}. This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem). 
As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g. \citep{korn1984ranges} or Fig. \ref{fig:sim-h0-demo} in this paper), as commonly seen in machine learning.

To overcome these limitations, I extend the permutation-based conditional independence framework of \cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output. 
The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).

The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models. 
First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
Second, functional connectivity data from the ABIDE \citep{di2014autism} database is used to investigate the potential motion- and center-bias \cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%TC:ignore
\section{Methods}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

@@

The proposed \emph{partial} and \emph{full confoudner tests} for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}. The inner workings of the tests is summarized on Fig. \ref{fig:overview}. In short, the tests model the conditional distribution between the confounder and the target variable with a generalized additive model (GAM) and then use a parallel-pairwise Markov-chain Monte-Carlo sampler that draws permutations of the confounder or target variable, for the partial and full confounder test, respectively. As a result, the permuted variables comply with the GAM-based conditional distribution. The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. The strength of this approach lies in avoiding to place any assumptions on the conditional distributions of the model output. This property ensures valid model diagnostics even in cases where Pearson and Spearman partial correlation and many other methods fail (as shown on Fig. \ref{fig:sim-h0-demo}).

As suggested by theory and shown by the simulations, both of the proposed tests provide a valid Type I error control (Fig. \ref{fig:sim-normal}), even in case of non-linearity and non-normality (Figs. \ref{fig:sim-h0-demo} and \ref{fig:sim-non-normal}), except when non-normality is very extreme (purple distribution on Fig. \ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6). With this the proposed tests overcome the limitations of partial Spearman correlation and linear conditional permutations, demonstrated on Fig. \ref{fig:sim-h0-demo}.

For detailed results, see also Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full} and \ref{fig:sim-bbb-sig-full}.

Power estimates for the 'partial' test (with normal and linear simulations) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%. With a sample size of 500, the same confounder bias is still detected with a power greater than 84\% in all of the simulation cases. A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%. Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 25\% of variance, it is almost certainly detected even with the low sample size of $n \geq 50$.

Simulations with non-normal and non-linear dependencies has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}).

\subsection{Neuroimaging data}
\subsubsection*{HCP dataset}

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. The problem of testing confounder-bias in predictive models can considered as assessing \emph{conditional independence} between the predicted values, the the target and the confounder variable.
Importantly, in predictive modelling, the model predictions are often characterized by a non-normal and/or non-linear conditional dependence on the target variable, thus any test for confounder bias must be, at least partly, non-parametric.
However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power. In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios. This is clearly demonstrated on Fig. \ref{fig:sim-h0-demo} in case of partial Spearman correlation. Although warnings about these issues were given from early on \citep{korn1984ranges} and received a fair amount of attention recently \citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated within the predictive modelling community.

From the three involved variables, the most problematic one is clearly the prediction output, known to often display non-normal and non-linear conditional dependencies on the other two. Therefore, a method being distribution-free for this variable, but inputting custom distributional information about the other two variables might provide a sufficient flexibility required for predictive model diagnostics.

To achieve this, here I propose to combine the conditional permutation test (CPT) of \cite{berrett2020conditional} with generalized additive model (GAM)-based conditional distribution modelling \citep{hastie1987generalized}. The proposed mathematical framework offers two novel tests for probing confounder bias: the partial and full confounder tests. These tests can handle both numerical and categorical data and can be applied for any predictive model, without having to re-fit the model, that is, with a negligible extra computational cost.

As expected from theory, both tests displayed a valid type I error control in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable.
Even though the tests' valid type I error control is justified by theory and reinforced by the simulations, a low statistical power could significantly limit their usefulness. The simulations presented in this paper revealed that both tests have a high, practically relevant power in a wide variety of scenarios. Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining about 4\% of the variance in the prediction). A lower sample size of 100 will still very likely identify medium sized confounder-biases (around 12\%). Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50. As expected, the power of the full confounder test is very similar to that of the partial confounder test, since the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.
While different biomedical applications might consider different amounts of bias to be relevant, an a-priori power calculation can help identifying the necessary sample size for proper model diagnostics.

>>>>

Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized, spisak2020pain, walsh2021dome}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Biomedical validity (i.e. whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers \citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases \citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability. Examples include models picking up on measurement artifacts (in-scanner head motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's \citep{rao2017predictive}, attention deficit hyperactivity disorder \citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD) \citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence \citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance \citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (racially biased machine learning models \citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.

While various data cleaning methods might help in mitigating confounder bias \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive, johnson2007adjusting}, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.

Powerful and robust statistical tests to quantify confounding effects in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches. Although confounder bias can be expressed in terms of conditional independence of the model predictions on the target and the confoudner variables, the proper evaluation of the conditional independence among these variables is challenging. Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation) are not valid measures of conditional independence. Although warnings about this issue were given from early on \citep{korn1984ranges}, and received a fair amount of attention recently \citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g. feature-set characteristics and model regularization \citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.

Alternative techniques for quantifying confoudner bias either fail to control type error (as  the known in the case of balanced permutations \cite{southworth2009properties, hemerik2018exact}, used in \cite{chaibub2019permutation}), or do not aim at all to do so \cite{ferrari2020measuring, wachinger2021detect}. 
Moreover, without some modifications, they are applicable for either only numeric \cite{wachinger2021detect} or only categorical variables \citep{chaibub2019permutation, ferrari2020measuring} and involve re-fitting the model \citep{chaibub2019permutation, ferrari2020measuring}, which might not be feasible for models with high a computational cost (e.g. with nested cross-validation). Some techniques \cite{wachinger2021detect}, furthermore, probe a null hypothesis (direct causation between the features and the prediction target) that is not applicable for for many kinds of data and models \footnote{For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation. In such cases, the alternative hypothesis of this test might be true in absence of any real confounder. }.

This work aims to construct a test of confounder bias which (i) can be applied post-hoc, without re-fitting the model (i.e. requires only the target variable, the confoudner and the model predictions), (ii) guarantees valid type-I error control even if the involved conditional distributions are non-normal and/or non-linear, (iii) is applicable for classification and prediction problems, both with numerical and categorical confounders.

@@

%In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}. This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem). 
%As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g. \citep{korn1984ranges} or Fig. \ref{fig:sim-h0-demo} in this paper), as commonly seen in machine learning.

%To overcome these limitations, I extend the permutation-based conditional independence framework of \cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output. 
%The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
%Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
%The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).

%The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models. 
%First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
%Second, functional connectivity data from the ABIDE \citep{di2014autism} database is used to investigate the potential motion- and center-bias \cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.

\section{Results}

The idea of conditional independence provides a straightforward framework for assessing the confounder bias in predictive models. However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models \citep{garcia2009study, kristensen2017whole} poses a great challenge.
In fact, as recently shown by Shah and colleagues \cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power. In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios. For instance, a trivial candidate, partial Spearman correlation, exhibits inflated type I errors even with slight violations of normality and linearity, as clearly demonstrated with simulated data on Fig. \ref{fig:sim-h0-demo}.

As the most problematic from the three variables (target, prediction, confounder) is clearly the model prediction \citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics. Exactly this can be achieved with the combination of the novel framework of conditional permutation testing (CPT) \cite{berrett2020conditional} and the use of generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}-based conditional distribution estimation. The proposed approach offers two novel tests for probing confounder bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.

The inner workings of the \emph{partial confoudner test} is summarized on Fig. \ref{fig:overview}. In short, the test models the conditional distribution between the confounder and the target variable with a generalized additive model (GAM) - or with a multinomial logistic regression, in case of categorical confounder - and then uses a so-called parallel-pairwise Markov-chain Monte-Carlo sampler that draws permutations of the confounder, so that the permuted variables still comply with the GAM-based conditional distribution. The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. The \emph{full confoudner test} works in an analogous way, with the difference that it creates permuted copies of the target variable, instead of the confoudner.

These tests can handle both numerical and categorical data and can be applied for any predictive model, without having to re-fit the model, that is, with a negligible extra computational cost.
The proposed \emph{partial} and \emph{full confoudner tests} have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.

As suggested by theory (see \hyperref[sec:methods]{Methods} for details) and shown by the simulations with a wide range of settings, both of the proposed tests provide a valid Type I error control (Fig. \ref{fig:sim-normal}), even in case of non-linearity and non-normality (Figs. \ref{fig:sim-h0-demo} and \ref{fig:sim-non-normal}), except when non-normality is very extreme (purple distribution on Fig. \ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6).

Estimates of statistical power for the partial confounder test (with normal and linear simulations, for  a wide range of parameters) are shown on Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%. With a sample size of 500, the same confounder bias is still detected with a power greater than 84\% in all of the simulation cases. A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%. Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confounder explains more than 25\% of variance, it is almost certainly detected even with the low sample size of $n \geq 50$.

Simulations with non-normal and non-linear dependencies has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}).
Type I error and power was found to be highly similar in case of categorical variables and for the \emph{full} confounder test, as well (Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full} and \ref{fig:sim-bbb-sig-full}).

\subsection{Neuroimaging data}

To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, they have been deployed in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.

\subsubsection*{HCP dataset}

Functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.

Functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

The assessment of confoudner bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings. However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity \citep{korn1984ranges} (see an example on Fig. \ref{fig:sim-h0-demo}). While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e. the nonexistence of a uniformly valid conditional independence test) \cite{shah2020hardness} clearly implies that conditional independence-based confoudner bias testing must be designed so that its suitability for the particular problem may be judged easily.

In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confoudner problem' in predictive modelling. Namely, the proposed approach places assumptions only about the conditional distribution of the confoudner and the target variables, as required by the applied generalized additive \cite{hastie1987generalized} or multinomial logistic model \cite{bennett1966multiple, jones1975proability}. In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analysis, as well. Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and data acquisition is probably already designed so that the assumptions are likely fulfilled.
Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization) \citep{garcia2009study, kristensen2017whole}. 
This property distinguishes the approach from other alternatives as it guarantees a valid type-I error control even in cases of non-normally and non-linearly dependent predictions, i.e. in cases where Pearson and Spearman partial correlations, and many other methods fail.

The proposed approach gives rise to two different statistical tests for testing confoudner bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable. 
The tests can be applied for arbitrary classification or regression models, without having to re-fit the model, that is, with a negligible extra computational cost.

As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except extreme non-normality), confirming the tests' can be deployed in most predictive modelling scenarios. While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculation in order to identify the necessary sample size for proper model diagnostics.
Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.

%In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g. \citep{spisak2020pain}. However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable. In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association. 
%Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
%These examples clearly indicate that guiding model diagnostics based on unconditioned associations between the confounder and the model output might result in overly strict decisions. More realistic is to aim at models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e. the prediction-confounder association is proportional to, and can be fully explained by, the target-confounder association.
%Put simply, this is what is tested by the proposed partial confounder test.

A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders typical for biomedical research (e.g. sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g. imaging artifacts) presents a great challenge to these efforts.
The usefulness of the proposed tests is demonstrated in two such examples, using the HCP \citep{van2013wu}  and the ABIDE \citep{di2014autism} datasets.
%First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
$Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
The hypothesised confounder effects were clearly present in both datasets.

In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings \citep{dubois2018distributed, lohmann2021predicting}) could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample. In this case, the bias would likely significantly harm the out-of-sample generalizability of this model. The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset. There can be manifold reasons for the observed bias. Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g. as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time). To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g. slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff). This is obviously undesirable for any brain-based predictive model of fluid intelligence.

In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD. For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion, e.g. due to atypical motor function (as known e.g. for ADHD \citep{eloyan2012automated}).

 The partial confounder test provided quantitative, statistically rigorous metrics for assessing the effectiveness of the investigated confounder mitigation techniques. In the HCP data, it revealed that both the acquisition bias and the age bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all). Given the high power of the test at the sample sizes of the HCP dataset ($N=999$), any remaining confounder-bias is most probably very safely negligible and well out of the range of practical relevance.

On the other hand, the partial confoudner test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset. The center bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT. Nevertheless, the p-values returned by the partial confounder test increased, in both cases, with two orders of magnitude as compared to the case with raw features, suggesting that a large part of center effects was still successfully removed by both approaches. Although determining the relevance of the remaining bias is out of the scope of this paper, significant p-values of the partial confounder test generally suggest that these models require more effective data harmonization.

Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign. Meanwhile, the prediction performance substantially dropped (although still remained statistically significant). 
A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
This effect might be boosted by the previously reported "residual motion artifacts" and their regional interactions \citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
 As COMBAT was originally developed for harmonizing effects of categorical variables (e.g. center or batch), its application for continuous confounder variables is not trivial. The presented method of inputting the binned "motion groups" into COMBAT resulted in an efficient removal of motion-bias and, at the same time, preserved a significant amount of signal-of-interest. 
 Nevertheless, inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.

 In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT and (iii) in some cases confoudner mitigation can - paradoxically - introduce more bias. The partial confounder test can be considered as a useful, objective benchmark to guide the search for a suitable confounder mitigation approach for every dataset.
 Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases. Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$). In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
This example highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.

To fill this critical gap in predictive model development, here I proposed two novel tests, the \emph{partial} and the \emph{full confounder tests}, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively. Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control and with assumptions that are easy to meet in most predictive modelling scenarios. The tests have a minimal computational overhead; as re-fitting the model is not required.

As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allow quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
Given their simplicity, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confoudner tests can become widespread tools that pave the way for the development of clinically useful machine learning biomarkers.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%TC:ignore
\section{Methods}
\label{sec:methods}

@@

@@

%TC:ignore 
\bibliographystyle{naturemag} %apalike}  
\bibliography{references}
